---
title: "Homework: Manipulating Data"
author: "Firstname Lastname"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3)
```


```{r load-sas-libraries, echo = F, message = F, warning = F}

library(SASmarkdown)
sas_enginesetup(sashtml=sashtml)

sasexe <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -ls 75"

# Linux options (for grading, please leave this in!)
# sasexe <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en"
# sasopts <- "-ls 75"
```

## Instructions

When you are finished with the assignment: 

1. Save the file as 06_Lastname_Firstname.Rmd and compile it
2. Commit the Rmd file and corresponding html file to your homework git repo
3. Push the commit to github
4. Locate the Rmd file on github and click on the SHA link for the file
5. Paste the url of the resulting page into Canvas to submit your assignment.

Your assignment must compile on a different computer as long as the saspath is set correctly for that machine. This means you will need to use a local file path when you read the data in via R (and SAS). 

## Poetry? The `R`aven
(In case it isn't obvious, do this part in R)

We can of course read in data from any number of sources, but we can also use R to analyze poetry and written text. Poetry is particularly interesting as it typically is stored in lines and stanzas, making it an ideal structured text to read in and create variables for. In this exercise, you'll be playing around with a poem and using string functions to explore the poem. We haven't covered string functions formally yet, so this is also a chance to explore the `stringr` package on your own (we'll formally talk about strings more next week). Think of this as good practice for reading the documentation of a package and learning how to leverage that documentation to accomplish a task.

> "The Raven" is a narrative poem by American writer Edgar Allan Poe. First published in January 1845, the poem is often noted for its musicality, stylized language, and supernatural atmosphere. Sitting on a bust of Pallas, the raven seems to further distress the protagonist with its constant repetition of the word "Nevermore". The poem makes use of folk, mythological, religious, and classical references. -- [Wikipedia](https://en.wikipedia.org/wiki/The_Raven)

Read in the lines of text in "The Raven" and write code to do the following tasks:

1. Create a tibble with a column that contains the lines of the poem (leave the blank lines in for now)
2. Count the number of words in each line. Do you see any patterns?
3. Can you use this information to create a variable with stanza numbers?
4. Create a column with line numbers (within each stanza)
5. Split the lines of the poem into words (you can discard punctuation). Is there any discernible pattern?

If this seems difficult, I understand. I have included a short demonstration of the assignment using the Edgar Allan Poe poem "Annabel Lee" so that you can get a sense of what I'm asking for. For each task, you may find it useful to do the task manually for a few lines, then write out a specific task list that summarizes the operations you had to do to complete the task. Use this list to create your sequence of steps by translating the steps into `dplyr` verbs.

If you are interested, you can use the `nsyllable` function in the R package `quanteda` to get syllable counts and continue the exploration. 

Hints: `stringr` is the package for text manipulation. `str_split` will split lines into words. You may also find `unnest` from the `tidyr` package to be useful 

You can get a preview of next week's material that covers `stringr` in more detail [here](https://srvanderplas.github.io/unl-stat850/transforming-data.html#string-operations-creating-new-variables-and-separating-multi-variable-columns)

## Superstitions?

Five Thirty Eight published a story called ["Some people are too superstitious to have a baby on Friday the 13th"](http://fivethirtyeight.com/features/some-people-are-too-superstitious-to-have-a-baby-on-friday-the-13th/), and, along with the story, they published the data to [this github repo](https://github.com/fivethirtyeight/data/tree/master/births). Download one of the CSVs in the repository and read it into SAS. 

1. Aggregate the data by month - is there any difference in the average number of births each month, across different months?

2. Aggregate the data by day of the week. Are there any days of the week that are less likely to have births?

3. Compute the day of the year (so Feb 1 would be 32, June 25 would be 176 if the year isn't a leap year, and 177 otherwise, and so on). What can you say about overall trends based on the day of the year? Are there any that would not be attributable to regular holidays, such as Memorial day, Labor Day, and Christmas? (Thanksgiving and Easter rotate around a bit more). Bonus points if you can figure out a macro that will compute this based on the year, month, and date.

4. Think through some way to numerically compare the births on Fridays (not the 13th) with births on Fridays that are the 13th. You can fit a model, compute means and subtract them -- whatever you think is appropriate. I will evaluate you on your code and logic, not on your statistical skills. 
